---
title: "ANOVA"
output: html_document
date: "2025-01-05"
---

# ANOVA (Analysis of Variance)

Used to compare means across more than two groups. ANOVA evaluates the variability within and between groups using the sum of squares.

#### Assumptions

-   Data are normally distributed within each group.
-   Groups have equal variances.

#### Hypotheses

-   **Null Hypothesis (H0)**: All group means are equal.
-   **Alternative Hypothesis (HA)**: At least one group mean is different.

#### Formula

$F = \frac{MS_{between}}{MS_{within}}$

#### Example Scenario

Does fertilizer type influence crop yield?

#### Data

```{r}
data <- data.frame(
  fertilizer = rep(c("Type A", "Type B", "Type C"), each = 10),
  yield = c(rnorm(10, mean = 20, sd = 3), rnorm(10, mean = 25, sd = 3), rnorm(10, mean = 22, sd = 3))
)
```

```{r anova-visualization}
boxplot(yield ~ fertilizer, data = data, main = "Yield by Fertilizer Type")
```

#### Code Example

```{r anova}

# Using aov (Type I)
anova_result <- aov(yield ~ fertilizer, data = data)
summary(anova_result)

# Using car package for Type II/III ANOVA
if (!requireNamespace("car", quietly = TRUE)) install.packages("car")
library(car)
Anova(lm(yield ~ fertilizer, data = data), type = "II")
```

#### Output Explanation

-   **F-statistic**: Ratio of variance between groups to variance within groups.
-   **p-value**: Tests the null hypothesis that all group means are equal.

#### Non-Parametric Alternative

Kruskal-Wallis test can be used for non-normal data.

```{r kruskal-wallis}
kruskal.test(yield ~ fertilizer, data = data)
```
