---
title: "Association between variables"
author: Vladislav Stanin  
date: 01/24/2024
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Two dichotomous variables (2 x 2 table)

|                                  | Cured | Dead | [*Row Total*]{.underline} |
|----------------------------------|-------|------|---------------------------|
| **Treatment**                    | 12    | 8    | 20                        |
| **Placebo**                      | 6     | 10   | 16                        |
| [***Column Total***]{.underline} | 18    | 18   | [**36**]{.underline}      |

```{r}
data_cure = data.frame(
  Patient_ID = 1:36,
  Group = factor(c(rep("Treatment", 12), rep("Placebo", 6), rep("Treatment", 8), rep("Placebo", 10))),
  Outcome = factor(rep(c('Cured', 'Dead'), e=18)))

data_cure %>%  head()
```

```{r}
contingency_table = table(data_cure$Group, data_cure$Outcome)
contingency_table
```

```{r}
contingency_table_big = addmargins(contingency_table)
contingency_table_big
```

## Risk ratio ($RR$)

-   $RR$ = 1 if two variables unassociated

$Risk_{tr}$ (to be dead in case of treatment) = $\dfrac{Dead  | Treatment}{Total\ Treatment}$

```{r}
risk_tr = contingency_table_big['Treatment', 'Dead']/ contingency_table_big['Treatment', 'Sum']
risk_tr
```

$Risk_{pl}$ (to be dead in case of Placebo) = $\dfrac{Dead  | Placebo}{Total\ Treatment}$ =

```{r}
risk_pl = contingency_table_big['Placebo', 'Dead']/ contingency_table_big['Placebo', 'Sum']
risk_pl
```

Risk ratio ($RR$) = $\dfrac{Risk_{tr}}{Risk_{pl}} =$

```{r}
RR = risk_pl / risk_tr
RR
```

There is some association between Treatment and Result - Placebo patients tend to be dead more often.

**If there were more cured:**

```{r}
risk_tr_1 = contingency_table_big['Treatment', 'Dead'] / 1020
risk_tr_1

risk_pl_1 = contingency_table_big['Placebo', 'Dead'] / 1016
risk_pl_1

RR = risk_pl_1 / risk_tr_1
RR

```

Almost the same $RR$, despite big number of cured!

## Risk difference ($RD$)

$RD = Risk_{tr} - Risk_{pl} =$

> *Use variables created above*

```{r}
# RD = 
# RD
```

**And if there are more Cured:**

```{r}
# RD = 
# RD
```

**-\> Always show both RR and RD in your text!**

## Odds ratio ($OR$)

$Odds_{tr} = \dfrac{Dead  | Treatment}{Cured | Treatment} =$

```{r}
# odds_tr = 
```

$Odds_{pl} = \dfrac{Dead  | Placebo}{Cured | Placebo} =$

```{r}
# odds_pl = 
```

$Odds \ ratio \ (OR) = \dfrac{Odds_{tr}}{Odds_{pl}} =$

```{r}
# OR = 
```

**And if there are more Cured:**

```{r}
# 
```

***better use OR than RR ...***

# Continuous x categorical

## Mean difference...

```{r}
data_md = data.frame(
  Meme_category = rep(c('A','B'), each=50) |> as.factor(),
  Laugh_level = c(rnorm(50, 8, 1), rnorm(50, 5, 1))
)

data_md |> glimpse(width = 50)
```

```{r}
mean_laughs = data_md |> group_by(Meme_category) |> summarise(mean_laugh = mean(Laugh_level))

MD = mean_laughs$mean_laugh[mean_laughs$Meme_category == 'A'] - 
  mean_laughs$mean_laugh[mean_laughs$Meme_category == 'B']

MD
```

# Continuous vs Continuous (Pearson's correlation)

**Pearson's correlation coefficient** measures the **strength** and **direction** of a linear relationship between two variables.

## Formula

-   Pearson's correlation coefficient:

    $\rho = \dfrac{\sum (X_i - \bar{X})(Y_i - \bar{Y})}{\sqrt{\sum (X_i - \bar{X})^2 \sum (Y_i - \bar{Y})^2}} \in [-1;1]$

### Interpretation

-   $−1$ indicates a perfect negative (inverse) linear relationship.

-   $+1$ indicates a perfect positive linear relationship.

-   $0$ indicates no linear relationship.

## Is there a relationship between temperature and metabolic rate?

### Data

```{r}
plant_height_df = tibble(
  Plant_ID = 1:28,
  # Sunlight_hrs_per_day = rep(4:10, e=4),
  Sunlight_hrs_per_day = runif(28, 4, 10),
  Plant_height = Sunlight_hrs_per_day + rnorm(28,2,3)
)

head(plant_height_df)
```

```{r}
plot(x = plant_height_df$Sunlight_hrs_per_day, 
     y=plant_height_df$Plant_height, 
     main = NULL, 
     xlab = "Sunlight (hrs/day)", 
     ylab = "Plant height", pch=19, col='blue')
```

### Calculation

-   `cor()` - calculates correlation matrix between all features of dataframe

```{r}
cor(plant_height_df$Sunlight_hrs_per_day, plant_height_df$Plant_height)
# cor(iris %>%select(where(is.numeric)))
```

-   `corr.test()` - also tests how far the coefficient is from zero *(t-test)*

```{r}
cor.test(plant_height_df$Sunlight_hrs_per_day, 
         plant_height_df$Plant_height)
```

## Example of 0 correlation

### 1

```{r}
x1=rnorm(500)
y1=rnorm(500)
plot(x1,y1)
```

```{r}
cor(x1, y1)
```

### 2

```{r}
x2=rnorm(500)
y2=x2^2 + rnorm(500, 0, 1)
plot(x2,y2)
```

```{r}
cor(x2, y2)
```

> **That's why you always should draw the scatterplot!**

### P.S

Game [**Guess the Correlation**](https://www.guessthecorrelation.com/)

# Ordinal vs ordinal \| continuous vs ordinal

## Math and Physics Grades

```{r}
grades_math_phys = data.frame(math = c(5,2,3,4,6,1,9,10,2,5),
                             phys = c(4,2,3,1,2,2,8,8,2,4))
grades_math_phys
```

## Rank correlations

> [The problems with Pearson correlation are that it is not suitable for "strange" non-linear relationships, non-normally distributed traits and data with outliers.]{.underline}

### Kendall's $\tau$

```{r}
cor(grades_math_phys, method = 'kendall')
```

### Spearman's $\rho$

-   Suitable for non-linear relationships

-   Less sensitive to outliers

-   Calculated based on ranks

```{r}
cor(grades_math_phys, method = 'spearman')
```

#### Examples

-   Non-linear

```{r}
x = runif(50, -10, 10)
y = (x + rnorm(50, 0, 2))^3
plot(x,y, pch=19)
```

```{r}
cor(x,y)
cor(x,y, method = "spearman")
# cor(x,y, method = "kendall")
```

-   With outliers

```{r}
x = c(rnorm(20,0,3), c(15, 19))
y = c(x[1:20] + c(rnorm(20,0,1)), 0, 0)
plot(x,y, pch=19)
```

```{r}
cor(x,y)
cor(x,y, method = "spearman")
# cor(x,y, method = "kendall")
```

# Conclusions

1.  **Correlation does not imply causation**
    -   Website with numerous [spurious correlations](https://tylervigen.com/spurious-correlations)
2.  **Correlation indicates presence of a relationship (association)**
3.  **Absence of correlation indicates absence of a linear relationship**
4.  **Absence of correlation [does not]{.underline} indicate absence of a relationship (association)**

# Practice

## Arthritis dataset

Reaction Velocity of an Enzymatic Reaction

```{r}
arth_df = read.csv('../data/Arthritis.csv', 
                   stringsAsFactors = T)

arth_df |> glimpse(width = 50)
```

```{r}
arth_df |> summary()
```

> **Is it more likely to have "Marked Imporved" if you are female?**
>
> -   dichotomous vs dichotomous

```{r}
# cont_table = 
```

```{r}
# 
```

> **Is there a significant difference in the age of women and men?**
>
> -   Continuous vs dichotomous

```{r}

```

## Diamonds dataset

```{r}
diamonds_small = diamonds |> slice_sample(n=1000)

diamonds_small |> glimpse(width = 50)
```

```{r}
diamonds_small |> summary()
```

> Are there some associated variables?
>
> -   continuous vs continuous
>
> -   ordinal vs ordinal (not forget -\> `as.numeric()`)

```{r}
# 
```

```{r}
# 
```
