---
title: "CLT"
author: Vladislav Stanin  
date: 01/24/2024
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)

theme_custom <- theme(
    axis.text = element_text(size = 15),
    axis.title = element_text(size = 20),
    plot.title = element_text(size = 20, hjust = 0.5),
    plot.subtitle = element_text(size = 15, hjust = 0.5),
    legend.title = element_text(size = 15),
    legend.text = element_text(size = 15),
    panel.background = element_rect(fill = "white"), 
    panel.grid = element_blank()
  )

theme_set(theme_custom)
```

# Distributions

### Normal distribution

```{r}
# here mean and sd - parameters in "general" population 
# n - number of elements we choose from population
norm_sample = rnorm(n=100, mean=0, sd=1)

#parameters of **sample**
mean(norm_sample)
sd(norm_sample)

```

### Exponential distribution

```{r}

exp_sample = rexp(n=1000, rate=2)

# mean()
# sd()
```

### Binomial distribution

```{r}

exp_sample = runif(n=1000, min=0, max=1)

# mean()
# sd()
```

# CLT

### Standard error

```{r}
se = function(x){
  sd(x)/sqrt(length(x))
}
```

```{r}
sample_size = 100 
n_samples = 10000

df_trial <- tibble(
  sample_ID = rep(1:n_samples, each=sample_size),
  value = runif(sample_size*n_samples, -10, 0)
  )

df_standart = 
  df_trial |> 
  group_by(sample_ID) |> 
  summarize(m = mean(value)) |> 
  mutate(z_m = (m - mean(m))/sd(m)) 

mean(df_standart$z_m)
sd(df_standart$z_m)

```

```{r}
 df_standart |> 
  ggplot() + 
  geom_density(aes(x=m),
               # bins=30, col='black',
               fill='skyblue') +
  geom_density(aes(x=z_m),
               # bins=30, col='black', 
               fill='red3')
```

### Sample from uniform distribution

```{r}
sample_size = 1000 

sample = runif(n=sample_size, min=0, max=10)

hist(sample)
```

```{r}
n_samples = 1000

df_trial <- data.frame(
  sample_ID = rep(1:n_samples, each=sample_size),
    value = runif(sample_size*n_samples, 0, 10) # generating sample_size*n_samples values from the same distibution
  )

df_trial |> group_by(sample_ID) |> 
  summarize(m = mean(value)) |> 
  ggplot() + geom_histogram(aes(x=m), bins=30, col='black', fill='skyblue')
```

### Sample from exponential distribution

```{r}
sample_size = 1000 

# sample = rexp()

# hist()
```

```{r, eval=F}
n_samples = 1000

df_trial <- data.frame(
  sample_ID = rep(1:n_samples, each=sample_size),
    # value = rexp()
  )

df_trial |>
  group_by(sample_ID) |> 
  # summarize(....) |> 
  
  ggplot() +
  geom_histogram(aes(x=m), bins=30, col='black', fill='skyblue')
```

# Confidence intervals

## Standard normal

```{r}
true_mean = 5
true_sd = 1
```
### Sample

```{r}
sample_size = 50

sample = rnorm(sample_size, true_mean, true_sd)

sample_mean = mean(sample)
# sample_sd = sd(sample_norm)
sample_se = se(sample)

hist(sample)
```

### CI (95%)

```{r}
quantiles = qnorm(0, 1, p = c(0.025, 0.975)) #+- 1.96

c(
  sample_mean + quantiles[1] * sample_se,
  sample_mean + quantiles[2] * sample_se
)
```

### Confidence level

```{r}
#| fig-height: 7
#| fig-width: 10

sample_size = 100 
n_samples = 1000
true_mean=5

df_trial <- tibble(
  sample_ID = rep(1:n_samples, each=sample_size),
  value = rnorm(sample_size*n_samples, true_mean, 2)
  )


quantiles = qnorm(0, 1, p = c(0.025, 0.975))


df_ci = df_trial |> group_by(sample_ID) |> 
  summarize(mu = mean(value), 
            se_sample = se(value), 
            conf.min = mu + quantiles[1] * se_sample,
            conf.max = mu + quantiles[2] * se_sample) |> 
  mutate(out = ifelse((true_mean < conf.max) & (true_mean > conf.min), 'ok', 'missed'))

# df_ci |>  
  # ggplot(aes(y=)) +
  # geom_errorbar(aes(xmin = , xmax = , col=)) +
  # geom_vline(aes(xintercept = ), col='red', lty=2, size=2) +
  # labs(y="Sample", color='CI covers true mean') 
  
  
```

**Try to change sample size and see what will happen!**

## t-distribution

```{r}
true_mean = 5
true_sd = 2

```
### Sample

```{r}
sample_size = 50

sample = rnorm(sample_size, true_mean, true_sd)

sample_mean = mean(sample)
# sample_sd = sd(sample_norm)
sample_se = se(sample)

hist(sample)
```

### CI (95%)

```{r}
quantiles = qt(df = length(sample)-1, p = c(0.025, 0.975)) 
quantiles
```
```{r}
c(
  sample_mean + quantiles[1] * sample_se,
  sample_mean + quantiles[2] * sample_se
)
```

**Try different size of sample!**

### Another method for calculation CI with t-distribution

Looking only on "*95 percent confidence interval:*"

```{r}
t.test(sample)
```

