---
title: "Intro into statistics (code)"
author: Vladislav Stanin  
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)

theme_custom <- theme(
    axis.text = element_text(size = 15),
    axis.title = element_text(size = 20),
    plot.title = element_text(size = 20, hjust = 0.5),
    plot.subtitle = element_text(size = 15, hjust = 0.5),
    legend.title = element_text(size = 15),
    legend.text = element_text(size = 15),
    panel.background = element_rect(fill = "white"), 
    panel.grid = element_blank()
  )

theme_set(theme_custom)
```

```{r}
sample_size = 100
```

# Distributions

### Normal distribution

```{r}
# here mean and sd - parameters in "general" population 
# n - number of elements we choose from population
norm_sample = rnorm(n=sample_size, mean=0, sd=1)

#parameters of **sample**
mean(norm_sample)
sd(norm_sample)

```

### Exponential distribution

```{r}
exp_sample = rexp(n=sample_size, rate=4)

mean(exp_sample)
sd(exp_sample)
```

### Uniform distribution

```{r}

unif_sample = runif(n=sample_size, min=0, max=1)

mean(unif_sample)
sd(unif_sample)
```

### Discrete (!) distributions

```{r}
rbinom(n=sample_size, size=10, prob=0.5)

```

```{r}
rpois(n = sample_size, lambda = 1)
```

## Quantiles

> Show **qua[rt]{.underline}iles** of `norm_sample`
>
> Show 2.5% and 97.5% **qua[nt]{.underline}iles** of `unif_sample`

```{r}
quantile(norm_sample)
quantile(unif_sample, pr=c(0.025, 0.975))
```

# Sampling

## From population defined by us

```{r}
true_mean = 10
true_sd = 2

# big_population = read_csv('data.csv')
big_population = rnorm(50000, mean = true_mean, sd = true_sd) |> round()

big_population |>  hist()
```

```{r}
sample_size = 10

sample_0 = sample(x=big_population, size=sample_size)
sample_0
```

## Setting the population by possible values and probabilities

**!NB sum of probabilities should be equal to 1**

```{r}
values = c(1,100,10000)
probs_of_values = c(0.2, 0.5, 0.3)

sum(probs_of_values) == 1
```

### Sampling

```{r}
sample_size = 10

sample_2 = sample(values, sample_size, replace=TRUE, prob=probs_of_values)
sample_2
```

-   `replace=TRUE` for repeatable drawing same values from possible values

## From mathematically defined distribution

```{r}
sample_size = 10

sample_1 = rnorm(sample_size, mean=0, sd=1)
sample_1
```

# CLT

### Standard error

```{r}
se = function(x){
  sd(x)/sqrt(length(x))
}
```

### Generating samples

```{r}

#uniform 
a=0
b=10
## -> mean = ??
population = runif(100000, a, b)

sample_size = 2
n_samples = 10000

df_trial <- tibble(
  sample_ID = rep(1:n_samples, each=sample_size),
  value = sample(x=population,
                 size=sample_size*n_samples, 
                 replace = TRUE,
                 prob=rep(1/length(population), length(population))
                 )
  )

df_trial
```

### Standardizing each sample

```{r}
df_standart = 
  df_trial |> 
  group_by(sample_ID) |> 
  summarize(m = mean(value)) |> 
  mutate(z_m = (m - mean(m))/sd(m) ) # <-- Formula! 

df_standart

```

```{r}
#for non-standard
mean(df_standart$m)
sd(df_standart$m)

#for standard
mean(df_standart$z_m)
sd(df_standart$z_m)
```

```{r}
 df_standart |> 
  ggplot() + 
  geom_density(aes(x=m),
               # bins=30, col='black',
               fill='skyblue') +
  geom_density(aes(x=z_m),
               # bins=30, col='black', 
               fill='red3')
```

### Sample from uniform distribution

```{r}
sample_size = 3

sample = runif(n=sample_size, min=0, max=10)

hist(sample)
```

### Generating samples

```{r}
n_samples = 1000

df_trial <- data.frame(
  sample_ID = rep(1:n_samples, each=sample_size),
    value = runif(sample_size*n_samples, 0, 10) 
  # generating sample_size*n_samples values from the same distibution
  )

df_means_from_samples = 
  df_trial |> group_by(sample_ID) |> 
  summarize(m = mean(value)) 

df_means_from_samples |> 
  ggplot() + 
  geom_histogram(aes(x=m), bins=30, col='black', fill='skyblue')
```

### Sample from exponential distribution

> Write the code!

```{r}
sample_size = 30 

sample = rexp(sample_size, rate=1)

hist(sample, breaks = seq(-1,8,0.1))
```

#### Generating samples

> Write the code!

```{r, eval=F}
n_samples = 1000

df_trial <- data.frame(
  sample_ID = rep(1:n_samples, each=sample_size),
    value = rexp(sample_size*n_samples, rate=2) 
  )
  
df_means_from_samples = 
  df_trial |> group_by(sample_ID) |> 
  summarize(m = mean(value)) 

df_means_from_samples |> 
  ggplot() + 
  geom_histogram(aes(x=m), bins=30, col='black', fill='skyblue')
```

# Confidence intervals

## Using standard normal

### Sample

```{r}
true_mean = 5
true_sd = 1

sample_size = 50

sample = rnorm(sample_size, true_mean, true_sd)

sample_mean = mean(sample)
# sample_sd = sd(sample_norm)
sample_se = se(sample)

hist(sample)
```

### CI (95%)

```{r}
quantiles = qnorm(0, 1, p = c(0.025, 0.975)) #+- 1.96

c(
  sample_mean + quantiles[1] * sample_se,
  sample_mean + quantiles[2] * sample_se
)
```

### Confidence level

```{r}

sample_size = 30 
n_samples = 1000
true_mean=1

df_trial <- tibble(
  sample_ID = rep(1:n_samples, each=sample_size),
  value = rexp(sample_size*n_samples, 1)
  )

quantiles = qnorm(0, 1, p = c(0.025, 0.975))


df_ci = df_trial |> group_by(sample_ID) |> 
  summarize(mu = mean(value), 
            se_sample = se(value), 
            conf.min = mu + quantiles[1] * se_sample,
            conf.max = mu + quantiles[2] * se_sample) |> 
  mutate(out = ifelse((true_mean < conf.max) & (true_mean > conf.min), 'ok', 'missed'))

  
```

```{r}
table(df_ci$out)
prop.table(table(df_ci$out))
```

```{r, dpi=100}
#| fig-height: 7
#| fig-width: 10
df_ci |>  
  ggplot(aes(y=sample_ID)) +
  geom_errorbar(aes(xmin = conf.min, xmax = conf.max, col=out)) +
  geom_vline(aes(xintercept = 1), col='red', lty=2, size=1) +
  labs(y="Sample", color='CI covers true mean') 
```

> **Try to change sample size and see what will happen!**

## t-distribution

```{r}
true_mean = 5
true_sd = 2

```

### Sample

```{r}
sample_size = 50

sample = rnorm(sample_size, true_mean, true_sd)

sample_mean = mean(sample)
# sample_sd = sd(sample_norm)
sample_se = se(sample)

hist(sample)
```

### CI (95%)

```{r}
quantiles = qt(df = length(sample)-1, p = c(0.025, 0.975)) 
quantiles
```

```{r}
c(
  sample_mean + quantiles[1] * sample_se,
  sample_mean + quantiles[2] * sample_se
)
```

### Confidence level

```{r}
sample_size = 10
n_samples = 1000
true_mean=5

df_trial <- tibble(
  sample_ID = rep(1:n_samples, each=sample_size),
  value = rnorm(mean=true_mean, sd=1, sample_size*n_samples)
  )

quantiles = qt(sample_size-1, p = c(0.025, 0.975))


df_ci_t = df_trial |> group_by(sample_ID) |> 
  summarize(mu = mean(value), 
            se_sample = se(value), 
            conf.min = mu + quantiles[1] * se_sample,
            conf.max = mu + quantiles[2] * se_sample) |> 
  mutate(out = ifelse((true_mean < conf.max) & (true_mean > conf.min), 'ok', 'missed'))
```

```{r}
table(df_ci_t$out)
prop.table(table(df_ci_t$out))
```

```{r, eval=FALSE}
#| fig-height: 7
#| fig-width: 10

df_ci_t |>  
  ggplot(aes(y=sample_ID)) +
  geom_errorbar(aes(xmin = conf.min, xmax = conf.max, col=out)) +
  geom_vline(aes(xintercept = 5), col='red', lty=2, size=1) +
  labs(y="Sample", color='CI covers true mean') 

```

**Try different size of sample!**

### Another method for calculation CI with t-distribution

Looking only on "*95 percent confidence interval:*"

```{r}
t.test(sample)
```
