---
title: "Homework #1: Exploring elephant data"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

Download the provided dataset `ElephantsMF.csv` and place it in your working directory.

The dataset contains the following columns:

\- `rownames`: Unique identifiers for each record (elephant).\
- `Age`: Age of the elephants (years).\
- `Height`: Height of the elephants (cm).\
- `Sex`: Sex of the elephants (`M` for Male, `F` for Female).

#### Instructions

> **Every result must be shown! If you create or chabging some variable/column/dataframe, show it!**

> Don't forget to save your changes where needed

For example, `dataset %>% function()` does not change the `dataset`, but `dataset = dataset %>% function()` does

------------------------------------------------------------------------

## Downloading libraries

```{r, warning=FALSE, message=FALSE}
library(readr)
library(dplyr)
library(tidyr)
```

# Tasks

## Vectors

#### 1. Create vector containing numbers from -20 to 50 and number 100

```{r}
# vec =
```

#### 2. Subtract by 10 and [then]{.underline} divide by 2 each value in this vector

```{r}
# 
```

## Data Loading

#### 3. Load the dataset using `read.csv()`, save it to variable `dataset` and show the first 10 rows.

```{r}
# dataset <- read...
#
```

-   Don't forget to set parameters to
    -   specify NAs values
    -   make character columns factors
    -   first column a rowname
-   `head(...)` - by default it returns first 5 rows, but we need 10 :)

**3.1 Make `dataset` a tibble**

```{r}
#
```

#### 4. Check the structure and summary of the data.

```{r}
#
```

-   *Just use special functions*

## Data Cleaning

#### 5. Check for missing values (`NA`) in the dataset (how many in each column?).

```{r}
#
```

#### 6. Verify that the `Sex` column contains only valid categories (`M` and `F`).

```{r}
#
```

-   *Tip*: just use `table()` or `unique()`

## Data Transformation

#### 7. Create a new column `Age_category` categorizing elephants based on `Age`:

-   "Young" for Age \< 5

-   "Sub-adult" for 5 \<= Age \< 15

-   "Adult" for Age \>= 15

```{r}
##
```

-   Use `cut()` - don't forget about parameter `labels = c(...)`
-   Ensure that you created column in your `dataset` !

#### 8. Create a column `Height_category` that splits `Height` into two categories: "Below Median" and "Above Median" based on the median height.

```{r}
#
```

-   `median()` and `ifelse()`
-   Don't forget to make new column a **factor**
-   Ensure that you created column in your `dataset` !

#### 9. Create a new column that contain the **Height in meters** (divide by 100), delete column with **Height in cm**.

```{r}
##
```

-   You can set a new name of column or leave the old one.

-   Remember how to delete column

-   *Tip*: `mutate()`. Parameter `.keep=...` may be useful to alternatively delete column

-   Ensure the column is created inside `dataset`

#### 10. Round column `Height` in meters to two decimal places (*e.g.* 1.4567 -\> 1.46), i.e. replace values in column with rounded ones 

```{r}
#
```

-   *Tip:* `round()` and look for a specific parameter

-   Ensure the column is modified inside `dataset`

## Exploratory Data Analysis

#### 11. Compute mean values of `Height` for each `Sex` *(in meters)*

```{r}
##
```

-   Don't forget that `mean()` works only on vectors

    -   `%>% pull(column)` -\> vector

    -   `data$column` -\> vector

-   *if you are advanced user, you can use `group_by()` and `summarize()`*

#### 12. Identify the row with the maximum `Height` and print this row (not number of row!).

```{r}
#
```

-   *Tip:* `which.max()`

#### 13. Show a counts and frequency tables for the `Age_category` column. Show both counts and frequencies

```{r}
##
```

-   Use `table()` and `prop.table()`

#### 14. Show a contingency table of `Sex` and `Height_category` *(number of elephants for every combination of these two categories)*

```{r}
#
```

-   Use `table()`

#### 15.**\*** Group dataset by `Height_category` and calculate `mean` and `sd` of `Age` for each group.

```{r}
#
```

-   use `group_by()` and `summarise()`

#### 16. Show the rows for elephants taller than 2 metres but shorter than 2.5 metres, and younger than 15 years old. Then select the `Age` and `Height` columns.

```{r}
#
```

-   use special `dplyr` functions

#### 17. Identify the top 3 tallest elephants and extract their `Age`, `Height`, and `Sex`.

```{r}
#
```

-   You may be using data arranging and `head()`

#### 18. Show rows of the 3 tallest and 3 the shortest elephants (not separately, but within one dataset)

```{r}
##

```

-   You may be using data arranging
-   basic functions (`head()` and `tail()`) can also be useful, but remember ***how*** ordered is your data
-   **+1 point if you make it by one line of code (show as one data.frame)**
    -   *Tip:* you can use either `n()` or `nrow()`
    -   *Tip:* you can use `slice()`

#### 19.\* Identify 3 oldest elephants for each `Sex` and extract their `Age` and `Height`.

```{r}
#
```

-   `group_by()` and `slice_head()`

## Exporting Data

#### 20. Save our modified `dataset` with new columns as a new CSV file.

```{r, eval=FALSE}
#
```

-   Name the file whatever you want

-   Set `row.names = FALSE` if you are using the base function. The function from `readr` works fine by itself

# Submission

Submit your `.Rmd` file and the final outputs (`csv` file)
