---
title: "Homework #1: Exploring elephant data"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

Download the provided dataset `ElephantsMF.csv` and place it in your working directory.

The dataset contains the following columns:

\- `rownames`: Unique identifiers for each record (elephant).\
- `Age`: Age of the elephants (years).\
- `Height`: Height of the elephants (cm).\
- `Sex`: Sex of the elephants (`M` for Male, `F` for Female).

<<<<<<< HEAD
### Instructions
=======
#### Instructions
>>>>>>> 52e1a05 (second_commit)

> **Every result must be shown! If you create or chabging some variable/column/dataframe, show it!**

> Don't forget to save your changes where needed

For example, `dataset %>% function()` does not change the `dataset`, but `dataset = dataset %>% function()` does

------------------------------------------------------------------------

## Downloading libraries

```{r, warning=FALSE, message=FALSE}
library(readr)
library(dplyr)
library(tidyr)
```

# Tasks

## 1. Data Loading

<<<<<<< HEAD
### 1. Load the dataset using `read.csv()`, save it to variable `dataset` and show the first 10 rows.
=======
#### 1. Load the dataset using `read.csv()`, save it to variable `dataset` and show the first 10 rows.
>>>>>>> 52e1a05 (second_commit)

```{r task-1}
# 
```

-   `head(...)` - by default it returns first 5 rows, but we need 10 :)

<<<<<<< HEAD
#### 1.1 Make `dataset` a tibble
=======
**1.1 Make `dataset` a tibble**
>>>>>>> 52e1a05 (second_commit)

```{r task-1.1}
# dataset = 
```

<<<<<<< HEAD
### 2. Check the structure and summary of the data.
=======
#### 2. Check the structure and summary of the data.
>>>>>>> 52e1a05 (second_commit)

```{r task-2}
# 
```

-   *Just use special functions*

## 2. Data Cleaning

<<<<<<< HEAD
### 3. Check for missing values in the dataset.
=======
#### 3. Check for missing values in the dataset.
>>>>>>> 52e1a05 (second_commit)

```{r task-3}
# 
```

-   *Just use special functions*

<<<<<<< HEAD
### 4. Verify that the `Sex` column contains only valid categories (`M` and `F`).
=======
#### 4. Verify that the `Sex` column contains only valid categories (`M` and `F`).
>>>>>>> 52e1a05 (second_commit)

```{r task-4}
# 
```

-   *Tip*: just use `table()` or `unique()`

## 3. Data Transformation

<<<<<<< HEAD
### 5. Transform the `Sex` column into a factor.
=======
#### 5. Transform the `Sex` column into a factor.
>>>>>>> 52e1a05 (second_commit)

```{r task-5}
# 
```

-   *Tip*: use `as.factor()`
-   Use function and ensure that you changed column in `dataset`

<<<<<<< HEAD
### 6. Create a new column `Age_category` categorizing elephants based on `Age`:
=======
#### 6. Create a new column `Age_category` categorizing elephants based on `Age`:
>>>>>>> 52e1a05 (second_commit)

```         
-   "Juvenile" for Age \< 5
-   "Sub-adult" for 5 \<= Age \< 15
-   "Adult" for Age \>= 15
```

```{r task-6}
# 
```

-   `cut()` - don't forget about parameter `labels = c(...)`

<<<<<<< HEAD
### 7. Create a column `Height_category` that categorizes `Height` into two categories: "Below Median" and "Above Median" based on the median height.
=======
#### 7. Create a column `Height_category` that categorizes `Height` into two categories: "Below Median" and "Above Median" based on the median height.
>>>>>>> 52e1a05 (second_commit)

```{r task-7}
# 
```

-   `mean()` and `ifelse()`
-   Don't forget to make new column a **factor**

## 4. Exploratory Data Analysis

<<<<<<< HEAD
### 8. Compute basic statistics (mean, median, standard deviation) for the `Age` and `Height` columns.
=======
#### 8. Compute basic statistics (mean, median, standard deviation) for the `Age` and `Height` columns.
>>>>>>> 52e1a05 (second_commit)

```{r task-8}
# 
```

<<<<<<< HEAD
### 9. Identify the row with the maximum `Height` and print it.
=======
#### 9. Identify the row with the maximum `Height` and print it.
>>>>>>> 52e1a05 (second_commit)

```{r task-9}
# 
```

-   *Tip:* `which.max()`

<<<<<<< HEAD
### 10. Show a counts and frequency tables for the `Age_category` column. Show both counts and frequencies
=======
#### 10. Show a counts and frequency tables for the `Age_category` column. Show both counts and frequencies
>>>>>>> 52e1a05 (second_commit)

```{r task-10}
# 
```

-   Use `table()` and `prop.table()`

<<<<<<< HEAD
### 11. Show a contingency table of `Age_category` and `Height_category` 
=======
#### 11. Show a contingency table of `Age_category` and `Height_category`
>>>>>>> 52e1a05 (second_commit)

```{r task-11}
# 
```

-   Use `table()`

<<<<<<< HEAD
### 12.**\*** Group dataset by `Height_category` and calculate `mean` and `sd` of `Age` for each group.
=======
#### 12.**\*** Group dataset by `Height_category` and calculate `mean` and `sd` of `Age` for each group.
>>>>>>> 52e1a05 (second_commit)

```{r task-12}
# 
```

-   use `group_by()` and `summarise()`

<<<<<<< HEAD
### 13. Show rows for elephants taller than the median height and select the `Age` and `Height` columns. 
=======
#### 13. Show rows for elephants taller than the median height and select the `Age` and `Height` columns.
>>>>>>> 52e1a05 (second_commit)

```{r task-13}
# 
```

-   use `dplyr` functions

<<<<<<< HEAD
### 14. Arrange dataset by `Height` and show the first and last 5 rows and show them.
=======
#### 14. Arrange dataset by `Height` and show the first and last 5 rows and show them.
>>>>>>> 52e1a05 (second_commit)

```{r task-14}
# 
```

-   *Tip:* you can use either `n()` or `nrow()`
-   *Tip:* `slice()`

## 5. Additional Transformations

<<<<<<< HEAD
### 15. Create a new column that contain the **Height in meters** (divide by 100), delete column with **Height in cm**.
=======
#### 15. Create a new column that contain the **Height in meters** (divide by 100), delete column with **Height in cm**.
>>>>>>> 52e1a05 (second_commit)

```{r task-15}
# 
```

-   You can set a new column name or leave the old one.

-   *Tip*: `mutate()`. Parameter `.keep=...` may be useful

-   Ensure the column is created inside `dataset`

<<<<<<< HEAD
### 16. Round Height in meters to two decimal places (*e.g.* 1.4567 -\> 1.46)
=======
#### 16. Round Height in meters to two decimal places (*e.g.* 1.4567 -\> 1.46)
>>>>>>> 52e1a05 (second_commit)

```{r task-16}
# 
```

-   *Tip:* `round()` and look for a specific parameter

-   Ensure the column is modified inside `dataset`

## 6. Summary Tasks

<<<<<<< HEAD
### 17. Identify the top 3 tallest elephants and extract their `Age`, `Height`, and `Sex`.
=======
#### 17. Identify the top 3 tallest elephants and extract their `Age`, `Height`, and `Sex`.
>>>>>>> 52e1a05 (second_commit)

```{r task-17}
# 
```

-   You may be using data arranging

<<<<<<< HEAD
### 18.**\*** Identify 3 oldest elephants for each `Sex` and extract their `Age`, `Height`.
=======
#### 18.**\*** Identify 3 oldest elephants for each `Sex` and extract their `Age`, `Height`.
>>>>>>> 52e1a05 (second_commit)

```{r task-18}
# 
```

## 7. Exporting Data

<<<<<<< HEAD
### 19. Save our modified dataset as a new CSV file. 
=======
#### 19. Save our modified dataset as a new CSV file.
>>>>>>> 52e1a05 (second_commit)

```{r task-19, eval=FALSE}
# 
```

-   Name the file whatever you want

-   Set `row.names = FALSE` if you are using the base function. The function from `readr` works fine by itself

# Submission

Submit your `.Rmd` file and the final outputs (csv dataset)
