---
title: "Homework #3"
output: html_document
---

```{r setup, echo=F}
knitr::opts_chunk$set(echo = T, warning = F, message = F,
                      fig.width = 7, fig.height = 5)
```

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
```

# Patients dataset

**Columns**

-   Age

-   Sex

-   Blood Pressure Levels (BP)

-   Cholesterol Levels

-   Na to Potassium Ratio

-   Glucose level

## 1. Dataset

### 1.1 Read the data from file `patients.csv`

-   [Don't forget about parameters for reading]{.underline}

```{r}
#
```

### 1.2 Show summary

```{r}
#
```

**1.2.1 Don't you see something strange?**

`<...WRITE HERE...>`

## 2. Cleaning

### 2.1. Deal with NAs

**2.1.1 Count number of NAs in each column**

```{r}
#
```

**2.1.2 Show** **rows with at least one NA in any column**

-   use `complete.cases()`

```{r}
#
```

**2.1.3 Do something** **with rows with NAs**

-   *Delete or remain ..?*

```{r}
#
```

### 2.2 Deal with outliers

**2.2.1 Which columns have outliers?**

-   *Create box-plots of continuous variables (columns)*

```{r}
#
```

`<...WRITE HERE what column have outliers ...>`

**2.2.2 [Filter]{.underline} out rows with obvious outliers in this column**

```{r}
#
```

## 3. EDA

### 3.1 Distributions of continuous data

-   Show histograms of `Age`, `Na_to_K`, `Glucose`

```{r}
#
```

### 3.2 Distribution of factor variables

**3.2.1 Draw a [barplot]{.underline} where the height represents the number of obseravtions in each blood pressure (`BP`) group and the `fill` represents `Sex`**

-   X-axis = `BP`, fill = `Sex`
-   "dodge" or "stack" barplot

```{r}
#
```

**3.2.2 Draw barplot where the height represents number of observations in each** `Cholesterol` **group and the `fill` represents blood pressure** (`BP`)

-   X-axis = `Cholesterol`, fill = `BP`
-   "dodge" or "stack" barplot

```{r}
#
```

*P.S. We will test the independence of `BP` and `Cholesterol` in 5.1*

## 4. Testing averages between 2 groups

### 4.1 Do average `Age`s significantly differ between 2 `Sex` groups?

**4.1.0 (\*) Draw the plot with two histograms of `Age`s for each `Sex` group**

```{r}
#
```

**4.1.1 Conduct the test**

-   Choose appropriate **test**

```{r}
#
```

**4.1.2 Write the conclusion**

`<...WRITE HERE...>`

### 4.2 Do average `Glucose` significantly differ between two `Cholesterol` groups?

**4.2.0 (\*) Draw the plot with two histograms of `Glucose` for each `Cholesterol` group**

```{r}
#
```

**4.2.1 Conduct the test**

-   Choose appropriate **test**

```{r}
#
```

**4.2.2 Write the conclusion**

`<...WRITE HERE...>`

## 5. Testing the independence between two categories

### 5.1 Do Blood pressure (`BP`) levels depend on `Cholesterol` levels?

**5.1.1 Create contingency table**

```{r}
#
```

**5.1.2 Conduct the test**

-   *Chi-square or Fisher exact?*

```{r}
#
```

**5.1.3 Write the conclusion**

`<...WRITE HERE...>`

## 6. Association

### 6.1 Do males have high `Cholesterol` more often than females? Calculate Risks, Odds, their ratios and Risk difference.

**6.1.1 Create contingency table**

```{r}
# cont_table = 
```

**6.1.2 Calculate Risks and Odds**

```{r}
# 
```

**6.1.3 Calculate RR, RD and OR**

```{r}
# RR = 
# RR

# RD = 
# RD

# OR = 
# OR
```

**6.1.4 Are there significant association between `Sex` and having high `Cholesterol` ?**

`<...WRITE HERE...>`

### 6.2 Are `Age` and `Glucose` correlated?

**6.2.0 (\*) Draw scatter-plot `Age` vs `Glucose`**

```{r}
#
```

**6.2.1 Show Pearson's correlation value**

-   Additionally **test** **how far the coefficient is from zero**

```{r}
#
```

**6.2.2 Is correlation is significant?**

`<...WRITE HERE...>`

## 7. Linear regression

### 7.1 Build a linear model of `Glucose` dependence on `Age`

**7.1.1 Build a model**

```{r}
#
```

**7.1.2 Show summary of the model**

```{r}
#
```

**7.1.3 Is coefficient significantly far from zero?**

`<...WRITE HERE...>`

**7.1.4 (\*) Plot the model line with scatte-rplot**

```{r}
#
```
